Influence of Connectivity on Dung Beetle Communities

### 2025-03-12 corridor_docs/eric_ms_thesis/MS_thesis.Rmd
thesis step-by-step

knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE)
library(here)
library(tidyverse)
library(kableExtra)
# library(gridExtra)
# library(egg)
# library(magick)
# library(citr)
# library(devtools)
# devtools::install_github("crsh/papaja")
```

@@

As human disturbances continue to expand into natural landscapes, intact habitats are becoming increasingly fragmented. This degradation lends to loss in biodiversity on a global scale and interruptions in ecosystem processes and functions [@haddad_habitat_2015]. Effects from isolation can vary, however as habitats are broken down community structures are significantly altered [@WOS:000419965700012]. Corridors have been shown to be an important mechanism for facilitating the movement of organisms through fragmented landscapes with the goal of minimizing negative consequences of fragmentation[@haddad_corridor_2003]. As disturbance continues to intensify, it is becoming increasingly more important to understand how different taxonomic groups. Here, we aim to gain an understanding of how dung beetles, a group of insects well known for strong dispersal ability in order to compete for ephemeral resources(Hanski and Cambefort 1991), interact with corridors in their landscapes.

Dung beetles have emerged as a model system with which to test spatial ecology hypotheses  [@ros_how_2012; @roslin_dung_2000].They are an incredibly well studied group of insects which are well known for driving a multitude of ecosystem functions (Hasan et al. 2024). The removal, breakdown, and burial of animal feces is an important ecosystem service provided by dung beetles such as enhanced nutrient cycling and soil quality and reduction of parasites on methane emissions from dung [@slade_disentangling_2016; @WOS:000350444300013]. Local assemblages of dung beetles can be species-rich with species comprising a broad range of functional traits (e.g., size, foraging style, resource-use) [@WOS:000891747700001]. Previous studies have shown that isolated patches of habitat frequently have lower dung beetle diversity and abundance than areas of continuous habitat, as well as documented their presence in linear strips of habitat that resemble corridors [@WOS:000707546700001].  Past studies have also focused on how landscape structure alters the community compositions of dung beetles[@WOS:000398662200001], yet large landscape scale experimental studies with carefully controlled and replicated treatments are non-existent for this model species.

Here, we  aim to determine how connectivity and fragmentation affect Species Richness and Diversity, Abundance, and functional diversity. We sampled dung beetle communities in experimental landscapes developed for the express purposes of comparing connected and isolated patches, as well as the effects of patch to edge ratio and distance to edge. To ask the question of (1) how landscape connectivity impacts dung beetle assemblages dung beetles were collected, identified, and counted with the expectation that biodiversity and abundance would be higher in patches connected by corridors. Additionally we asked (2) Are corridors benefiting any one functional trait over another? Since our experimental system consists of open habitats amongst a forested matrix, we anticipate that species preferring open areas and generalists may be more common in our sampling.

@@

All dung beetles were counted and identified to species with the exception of beetles of the genus Aphodius, overall 15 species were identified and approximately 5300 individual beetles were collected.

@@

- species richness

  > glmm with poisson dist reccomended by julian

- species diversity

  > glmm with poisson dist reccomended by julian

- functional diversity

  > 1st pass: roller, tunneler, dweller  
  > after all done: morphological measurements like Alonso et al 2022 (see table from decastro 2023, Giménez Gómez 2025)

- abundance (or biomass)

  > abundance of the most common species?  
  > abundance of species caught in all sites?  
  > changes in rank-abundance.

_Reason for this breakdown: _

- richness and diversity are "community level" response. 
- abundance is a species-level response. 
- changes in abundance or rank-abundance  could have implications for ecosystem services if it is connected to finctional group.

Biodiversity between patch types was compared using Hill numbers (Jost 2006). We looked at community composition by increasing magnitudes of diversity components (qD) of 0D (species richness), 1D (Shannon entropy), and 2D (Simpson Diversity). Diversity numbers were calculated in R studio using package iNEXT (Chao et al. 2016). Bray-Curtis dissimiolarity values were calculated using package Vegan in R studio. Dung beetles were assigned traits by waste removal guild and habitat preference.

Insert description above of individual trap

1. **Functional Diversity:** Need to assign each species to a functional group: roller, tunneler. dweller, others? 
  - habitat preference (forest, pasture, generalist)

@@

## Overview

- how many beetles total
- most common species overall
- was. the same species the most common one in all habitats?
- were there any that were caught in all habitats? Any restricted to only one?
- any invasive / exotic species?

Overall, `r total_N` total dung beetles were collected belonging to N = `r total_spp` species. The most dominant species were ______  (N = _____), Ateuchus lecontei (n=1115), and Phanaeus igneus (n=958) (Table 1). All but one species was found in matrix plots. Four species (Onthophagus striatulus, Onthophagus concinnus, Geotrupes Blackburnii, Onthophagus truberculifrons) were the only species not representative in every patch type. We had two singleton species, Geotrupes Blackburnii appearing only once in winged, and Onthophagus tuberculiofrons  in matrix.

Beetle abundance was significantly higher in matrix plots with 33% (n = 1713) of captures followed by 26% (n = 1359) caught in connected patches, 23% (n = 1199) in winged, and lastly 18% (n = 942) from rectangular patches (Table 2).

Species richness was distinctly different between patch types with matrix plots exhibited the highest richness while connected patches were lowest (Fig. 1).  Estimations of Shannon diversity index were fairly uniform between patch types. Simpson diversity indices were similarly even however winged patches showed a lower value.

Notable Bray-Curtis dissimilarity values were between connected and matrix (BC = 0.127), connected and winged (BC = 0.077),  matrix and rectangle (BC = 0.318), and matrix and winged (BC = 0.189) (Table 3).

1. total number of beetles _from_ total number of species 
1. were all species found in all habitats? Were any species found in only 1 habitat?
1. Were all species found in all blocks? Were any restricted to only 1-2 blocks?
1. Were all species found in the matrix? (expect that so, since it is the 'baseline' or 'source' habitat)
1. Number of species in each functional group
1. most common 3-4 species
1. any rare species?

@@

\newpage
# FIGURES

@@

![Experimental landscapes indicating location of the pitfall traps.](pitfalls.png)

```{r Figure_hill, echo=FALSE,message = FALSE,warning=FALSE}
library(iNEXT)
#
# btl_data<-read_csv(here("corridor_docs","eric_ms_thesis","ms_data","data_clean","clean_btl_counts.csv"))
# spp_codes<-read_csv(here("corridor_docs","eric_ms_thesis","ms_data","data_raw","species_codes.csv"))

btl_sums_patch<-btl_data %>%
  pivot_longer(pvin:osyl,names_to = "species",values_to = "n") %>%
  group_by(species,patch,block) %>%
  mutate(patch=case_when(
    patch == "c" ~ "Corridor",
    patch == "m" ~ "Matrix",
    patch == "w" ~ "Winged",
    patch == "r" ~ "Rectangle",
    .default = as.character(patch))) %>%
  summarize(n=sum(n, na.rm=TRUE)) %>%
  arrange(patch,desc(n)) %>%
  rename(sp_code=species,
         n_patch=n) %>%
  mutate(pb=paste(patch,block,sep="-")) %>% 
  ungroup() %>% 
  select(-patch,-block) %>% 
  pivot_wider(
    names_from = pb,
    values_from = c(n_patch))

div.data <- btl_sums_patch |>
  column_to_rownames(var = "sp_code")

div.btl <- iNEXT(div.data, q = c(0,1,2), datatype = "abundance")

z <- iNEXT(div.data, q=c(0,1,2), datatype="abundance")
ggiNEXT(z, facet.var="Order.q", color.var="Assemblage") +
  theme_bw(base_size=18)

@@

spp_table<-knitr::kable(spp_table_data, 
      digits = 2,
      align="lcrcccc",
      format="latex",
      row.names = FALSE,
      booktabs=T,
      linesep = "", #removes the blank line after every 5 lines
      caption = "Dung beetle species sampled in the SRS site and their total abundance over the course of the study.") %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = F,
                latex_options="scale_down",
                font_size = 12,
                position = "left") %>% 
  column_spec(1, italic = T) %>% 
  row_spec(0, bold = T) %>% 
  row_spec(1:15, hline_after = T) 
spp_table

@@

abund_patch <- knitr::kable(btl_sums_patch, 
      digits = 2,
      align="lr",
      format="latex",
      row.names = FALSE,
      booktabs=T,
      linesep = "", #removes the blank line after every 5 lines
      caption = "Total dung beetles captured in all replicates of a patch type.") %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = F,
                latex_options="scale_down",
                font_size = 12,
                position = "left") %>% 
  # column_spec(1, italic = T) %>% 
  row_spec(0, bold = T) 
  # row_spec(1:15, hline_after = T) 
abund_patch

@@

@@

taxa_bray <- vegan::vegdist(btl_spec_patch, method = "bray", binary = FALSE)
taxa_bray

```
\newpage

@@

### Dedication 
To my family who never stopped supporting me along this journey, my friends who kept me company along the way, and my mentors at VCU who believed in me before I did myself.

### List of Abbreviations

@@

# Biographical Sketch

>>>>

knitr::opts_chunk$set(echo = FALSE,message=FALSE,warning=FALSE)
library(here)
library(tidyverse)
library(kableExtra)
library(grateful)
# library(gridExtra)
# library(egg)
# library(magick)
# library(citr)
# library(devtools)
# devtools::install_github("crsh/papaja")
cite_packages(citation.style = "peerj", out.dir = ".")
```

@@

**_Note from EB:_** _I don't want to see the words 'dung beetles' until paragraph 5_.

Paragraph 1: Overall Topic & Why it is interesting and Important.

>Topic Sentence:

Paragraph 2: What we know.

>Topic Sentence:

Paragraph 3: What we don't know.

>Topic Sentence:

Paragraph 4: Why we don't know it. _(hint: ecosystem services)_

>Topic Sentence:

Paragraph 5: Here is what we did to address this unknown.

>Topic Sentence:

---

> As human disturbances continue to expand into natural landscapes, intact habitats are becoming increasingly fragmented. This degradation lends to loss in biodiversity on a global scale and interruptions in ecosystem processes and functions [@haddad_habitat_2015]. Effects from isolation can vary, however as habitats are broken down community structures are significantly altered [@WOS:000419965700012]. Corridors have been shown to be an important mechanism for facilitating the movement of organisms through fragmented landscapes with the goal of minimizing negative consequences of fragmentation[@haddad_corridor_2003]. As disturbance continues to intensify, it is becoming increasingly more important to understand how different taxonomic groups. Here, we aim to gain an understanding of how dung beetles, a group of insects well known for strong dispersal ability in order to compete for ephemeral resources(Hanski and Cambefort 1991), interact with corridors in their landscapes.

> Dung beetles have emerged as a model system with which to test spatial ecology hypotheses  [@ros_how_2012; @roslin_dung_2000].They are an incredibly well studied group of insects which are well known for driving a multitude of ecosystem functions (Hasan et al. 2024). The removal, breakdown, and burial of animal feces is an important ecosystem service provided by dung beetles such as enhanced nutrient cycling and soil quality and reduction of parasites on methane emissions from dung [@slade_disentangling_2016; @WOS:000350444300013]. Local assemblages of dung beetles can be species-rich with species comprising a broad range of functional traits (e.g., size, foraging style, resource-use) [@WOS:000891747700001]. Previous studies have shown that isolated patches of habitat frequently have lower dung beetle diversity and abundance than areas of continuous habitat, as well as documented their presence in linear strips of habitat that resemble corridors [@WOS:000707546700001].  Past studies have also focused on how landscape structure alters the community compositions of dung beetles[@WOS:000398662200001], yet large landscape scale experimental studies with carefully controlled and replicated treatments are non-existent for this model species.

> Here, we  aim to determine how connectivity and fragmentation affect Species Richness and Diversity, Abundance, and functional diversity. We sampled dung beetle communities in experimental landscapes developed for the express purposes of comparing connected and isolated patches, as well as the effects of patch to edge ratio and distance to edge. To ask the question of (1) how landscape connectivity impacts dung beetle assemblages dung beetles were collected, identified, and counted with the expectation that biodiversity and abundance would be higher in patches connected by corridors. Additionally we asked (2) Are corridors benefiting any one functional trait over another? Since our experimental system consists of open habitats amongst a forested matrix, we anticipate that species preferring open areas and generalists may be more common in our sampling.

@@

All dung beetles were counted and identified to species with the exception of beetles of the genus Aphodius, overall 15 species were identified and approximately 5300 individual beetles were collected. 
 - what guides/keys did you do to ID? 
 - did you deposit vouchers and if so where are they available?

@@

### Overview

1. how many beetles total
1. most common species overall. 
1. was. the same species the most common one in all habitats?
1. were there any that were caught in all habitats? Any restricted to only one?
1. any invasive / exotic species?
1. Table: count by species by habitat

### Species richness

1. glmm with poisson dist reccomended by julian
1. table: results of analyses. 
1. visualization: hill's plots? rank-abundance plots? rarefaction curve?

### Species diversity

1. glmm with poisson dist reccomended by julian
1. table: results of analyses
1. visualization: hill's plots? rarefaction curve?

### Functional diversity

1. 1nd pass: roller, tunneler, dweller
1. 2nd: morphological measurements like Alonso et al 2022 (see table from decastro 2023, Giménez Gómez 2025)
1. table: results of analyses 
1. visualization: biomass of different (R/T/D) functional groups?

_Reason for this breakdown: _ - richness and diversity are "community level" response. abundance is a species-level response. changes in abundance or rank-abundance  could have implications for ecosystem services if it is connected to functional group.

---

> Biodiversity between patch types was compared using Hill numbers (Jost 2006). We looked at community composition by increasing magnitudes of diversity components (qD) of 0D (species richness), 1D (Shannon entropy), and 2D (Simpson Diversity). Diversity numbers were calculated in R studio using package iNEXT (Chao et al. 2016). Bray-Curtis dissimiolarity values were calculated using package Vegan in R studio. Dung beetles were assigned traits by waste removal guild and habitat preference.

> Functional Diversity: Need to assign each species to a functional group: roller, tunneler. dweller, others?  - habitat preference (forest, pasture, generalist)
 - what stats / computer packages did you use to analyze and visualize the data?

@@

### Overview

1. total number of beetles _from_ total number of species 
1. were all species found in all habitats? Were any species found in only 1 habitat?
1. Were all species found in all blocks? Were any restricted to only 1-2 blocks?
1. Were all species found in the matrix? (expect that so, since it is the 'baseline' or 'source' habitat)
1. Number of species in each functional group
1. most common 3-4 species
1. any rare species? 
1. any invasive / exotic species?

### Species richness

### Species diversity

### Functional diversity

---

> Overall, `r total_N` total dung beetles were collected belonging to N = `r total_spp` species. The most dominant species were ______  (N = _____), _Ateuchus lecontei_ (n=_____), and _Phanaeus igneus_ (n=_958___) (Table 1). All but one species was found in matrix plots. Four species (_Onthophagus striatulus_, _Onthophagus concinnus_, _Geotrupes blackburnii_, _Onthophagus truberculifrons_) were the only species not representative in every patch type. We had two singleton species, _Geotrupes blackburnii_ appearing only once in winged, and _Onthophagus tuberculiofrons_ in matrix.

> Beetle abundance was significantly higher in matrix plots with 33% (n = 1713) of captures followed by 26% (n = 1359) caught in connected patches, 23% (n = 1199) in winged, and lastly 18% (n = 942) from rectangular patches (Table 2).

> Species richness was distinctly different between patch types with matrix plots exhibited the highest richness while connected patches were lowest (Fig. 1).  Estimations of Shannon diversity index were fairly uniform between patch types. Simpson diversity indices were similarly even however winged patches showed a lower value.

> Notable Bray-Curtis dissimilarity values were between connected and matrix (BC = 0.127), connected and winged (BC = 0.077),  matrix and rectangle (BC = 0.318), and matrix and winged (BC = 0.189) (Table 3).

@@

\newpage
# FIGURES & TABLES

1. ~~Table: count by species by habitat.~~
1. table: results of species richness analyses.
1. Figure of species richness results.
1. table: results of species diversity analyses.
1. Figure of species diversity results.
1. table: results of functional diversity analyses.
1. Figure of biomass of different (R/T/D) functional groups.
1. Map of where field site is located.
1. ~~Figure of the patch layout with pitfalls~~.
1. Pictures of different species, sampling, field sites?

@@

![Experimental landscapes indicating location of the pitfall traps.](pitfalls.png){height=50%}

<!-- ```{r Figure_hill, echo=FALSE,message = FALSE,warning=FALSE} -->
<!-- library(iNEXT) -->
<!-- # -->
<!-- # btl_data<-read_csv(here("corridor_docs","eric_ms_thesis","ms_data","data_clean","clean_btl_counts.csv")) -->
<!-- # spp_codes<-read_csv(here("corridor_docs","eric_ms_thesis","ms_data","data_raw","species_codes.csv")) -->

<!-- btl_sums_patch<-btl_data %>% -->
<!--   pivot_longer(pvin:osyl,names_to = "species",values_to = "n") %>% -->
<!--   group_by(species,patch,block) %>% -->
<!--   mutate(patch=case_when( -->
<!--     patch == "c" ~ "Corridor", -->
<!--     patch == "m" ~ "Matrix", -->
<!--     patch == "w" ~ "Winged", -->
<!--     patch == "r" ~ "Rectangle", -->
<!--     .default = as.character(patch))) %>% -->
<!--   summarize(n=sum(n, na.rm=TRUE)) %>% -->
<!--   arrange(patch,desc(n)) %>% -->
<!--   rename(sp_code=species, -->
<!--          n_patch=n) %>% -->
<!--   mutate(pb=paste(patch,block,sep="-")) %>%  -->
<!--   ungroup() %>%  -->
<!--   select(-patch,-block) %>%  -->
<!--   pivot_wider( -->
<!--     names_from = pb, -->
<!--     values_from = c(n_patch))  -->

<!-- div.data <- btl_sums_patch |> -->
<!--   column_to_rownames(var = "sp_code") -->

<!-- div.btl <- iNEXT(div.data, q = c(0,1,2), datatype = "abundance") -->

<!-- z <- iNEXT(div.data, q=c(0,1,2), datatype="abundance") -->
<!-- ggiNEXT(z, facet.var="Order.q", color.var="Assemblage") + -->
<!--   theme_bw(base_size=18) -->

<!-- ``` -->

@@

spp_table<-knitr::kable(spp_table_data, 
      digits = 2,
      align="lcrcccc",
      format="latex",
      row.names = FALSE,
      booktabs=T,
      linesep = "", #removes the blank line after every 5 lines
      caption = "Dung beetle species sampled in the SRS site and their total abundance over the course of the study.") %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = F,
                latex_options="scale_down",
                font_size = 12,
                position = "center") %>% 
  column_spec(1, italic = T) %>% 
  row_spec(0, bold = T) %>% 
  row_spec(1:15, hline_after = T) 
spp_table

```
---

@@

abund_patch <- knitr::kable(btl_sums_patch,
      digits = 2,
      align="lr",
      format="latex",
      row.names = FALSE,
      booktabs=T,
      linesep = "", #removes the blank line after every 5 lines
      caption = "Total dung beetles captured in all replicates of a patch type.") %>%
  kable_styling(bootstrap_options = c("striped"),
                full_width = F,
                latex_options="scale_down",
                font_size = 12,
                position = "center") %>%
  # column_spec(1, italic = T) %>%
  row_spec(0, bold = T)
  # row_spec(1:15, hline_after = T)
abund_patch

@@

taxa_bray <- vegan::vegdist(btl_spec_patch, method = "bray", binary = FALSE)
# taxa_bray

```
---

@@

@@

## Dedication 
To my family who never stopped supporting me along this journey, my friends who kept me company along the way, and my mentors at VCU who believed in me before I did myself.

## List of Abbreviations

@@

## Biographical Sketch
